### 2019-08-14 Writeup/Thesis/pages/introduction.tex
Changes; - Corrections and wording changes

The increasing use of public key cryptography by instant messaging and secure email means key fingerprint  verification is an ever more important task.

One of the most significant risks to the security of the communication channel is a Man-in-the-middle (MiTM) attack. A MiTM attack involves an attacker impersonating one or both sides of a connection. MiTM attacks can fully circumvent the encryption as it allows an attacker to read all of the encrypted data.

A countermeasure for this is the verification of each parties' key fingerprint. A fingerprint is a small string of characters that is unique to each key pair. Therefore, it can be used as a way of identification.

Fingerprints can come in a number of different encodings such as hexadecimal, words and even procedurally generated avatars. Previous research has shown that the average human can only hold around 7-digits worth of data in their working memory\cite{miller1956magical}, alongside this the involvement of human based verification makes the designing of user friendly schemes a task of upmost importance.

Humans have long been known as the most vulnerable part of any system. To combat this issue, solutions have been proposed to remove the manual verification with examples such as PGP's web-of-trust\cite{callas1998openpgp} and Namecoin\cite{kalodner2015empirical}. However, these suffer from user adoption due to perceived complexity. This leaves manual verification in a difficult position, due the proposed solutions needing to sacrifice either usability or security. Therefore, research into improving or creating secure fingerprint encodings that work well with users remains massively relevant.

The report will asses the security and feasibility of attacking a fingerprint encoding known as ``Trustwords'' as justification for various design choices have been unsubstantiated.

The report will be structured as follows:

\textbf{Chapter 2} provides any required background knowledge alongside a review on all relevant literature.

@@

\textbf{Chapter 7} evaluates the success of the project alongside a discussion into relevant further work. The research aims proposed earlier in Chapter 3 will be used as evaluative metrics.

\textbf{Chapter 8} concludes the project with a culmination of the papers findings alongside any conclusions that can be drawn from the research.

>>>>

The increasing use of public-key cryptography by instant messaging and secure email means key-fingerprint  verification is an ever more important task.

One of the most significant risks to the security of the communication channel is a Man-in-the-middle (MiTM) attack. A MiTM attack involves an attacker impersonating one or both sides of a connection. MiTM attacks can entirely circumvent the encryption as it allows an attacker to read all of the encrypted data.

A countermeasure for the threat of MiTM attacks is the verification of each parties’ fingerprint. A fingerprint is a small string of characters that is unique to each key and, thus, can be used to identify.

Fingerprints can come in several different encodings such as Hexadecimal, words, and even procedurally generated avatars. Previous research has shown that the average human can only hold around 7-digits ($\pm 2$) worth of data in their working memory\cite{miller1956magical}. Consequently, this makes the designing of user-friendly schemes a task of utmost importance.

Humans are commonly considered the most vulnerable part of any computer system. Solutions, therefore, have been proposed to remove this manual verification with examples such as PGP's web-of-trust\cite{callas1998openpgp} and Namecoin\cite{kalodner2015empirical}. However, these suffer from user adoption due to perceived complexity. Manual verification is, consequently, left in a difficult position, due in part to proposed solutions needing to sacrifice either usability or security. Therefore, research into improving or creating a more secure and user-friendly fingerprint encoding remains an important task.

This report evaluates the security and feasibility of attacking a fingerprint encoding known as “Trustwords.” The motivation for this is due in part to the shortage of justification behind some fundamental design choices.

The report is structured as follows:

\textbf{Chapter 2} provides any required background knowledge alongside a review of all relevant literature.

@@

\textbf{Chapter 7} evaluates the success of the project alongside a discussion into proposed further work. The research aims proposed earlier in Chapter 3 are used as evaluative metrics.

\textbf{Chapter 8} concludes the project with a culmination of the findings of the paper alongside any relevant conclusions.

### 2019-08-14 Writeup/Thesis/pages/lit_review.tex
Changes; - Corrections and wording changes

The chapter will contain the following sections:

\begin{itemize}
    \item \textbf{Project context} \\ 
    Will provide the user with the base knowledge required to understand the context and purpose of the following literature review.

    \item \textbf{Definition of terms}\\
    Specific definitions relevant to the literature review will initially be defined.

    \item \textbf{Literature Review} \\
    A review of currently available literature on relevant topics. Recommendations and research gaps will be identified and discussed.

@@

To fully understand the literature review it is necessary to introduce and explain the base knowledge before continuing.

@@

Asymmetric cryptography facilitates the secure encryption of messages in end-to-end encryption (E2EE), verification of digital signatures and sharing of pre-communication secrets, among others.

The asymmetry stems from the use of "Public" and "Private" keys. The Public key is used to encrypt data that only the respective Private key can decrypt. This means the keys required to encrypt can be easily shared across insecure channels.

To construct an E2EE connection the initial stage will be a key exchange using the key types discussed previously. This is the sharing of a pre-shared secret between two verified parties, this will then be used to encrypt subsequent messages with symmetric encryption\footnote{This is due to the speed increase of symmetrical over asymmetrical encryption}. This, however, hinges on the verification of the initial parties, if one party impersonates another and gets to view the pre-shared secret all communication becomes decryptable. Therefore, the correct identification of parties is crucial to maintaining secure communication channels. The exploitation of this is known as a Man-in-the-middle (MiTM) attack. This can be bidirectional where the attacker can sit in the middle of a communication channel and decrypt all correspondence. Figure \ref{fig:mitm} contains a visual representation of this attack.

@@

Due to the assumption that the attacking party (Eve) does not have the same public-private key pair as either party (Alice or Bob) the fingerprint of the public key can be used for identification. There are, however, issues in how a user should link a key's fingerprint to a real world entity. This paper will, however, assume that users have access to the respective parties' correct fingerprint, as the protocols used to achieve this are outside the project's scope.

The fingerprint of the key is generated by running the main key components through a secure one way hash function such as SHA-1. This process produces a digest of fixed length that can be used to compare keys. Therefore, the comparison of expected and actual fingerprints can be used to detect MiTM attacks.
Historically these have been represented as a hexadecimal string whereon verification fingerprints are compared between two substrates, for example, a monitor screen and a business card.
This process is known as the \textit{``authentication ceremony''}. However, system designers are moving away from this structure and are now creating a combined key from each sides' public key. In the case of WhatsApp for example, the connection fingerprint of two parties keys is the first 30 bytes of a SHA-512\footnote{5200 iterations} hash of each parties' identity key, these are then both concatenated together to form a single key\cite{whatsapp2017paper}. This produces a unique key communication pair.
\\\\
Due to the manual comparison of the fingerprint, length and format is a consideration. Prior research has shown that the average human can only hold around 7-digits worth of data in their working memory\cite{miller1956magical}. This rules out the possibility of comparing complete digests. For example, SHA-1 is 40 hex digits (160-bit) and, therefore, difficult to effectively compare. Therefore, if human interaction is required; there is a need for schemes that work effectively with consideration to individual limitations.

Research in this area has taken various encoding schemes and compared their fallibility to impersonated attacks. The main elements of comparison have been the ``accuracy of attack detection" and ``time to compare". In this paper, these are considered the metrics of ``security'' and ``usability'', respectively.

\section{Definition of terms}
Throughout the paper, various terms are used to define elements of the authentication ceremony relevant to fingerprints verification. Explicit descriptions are, therefore, stated in this section. These are used to remove ambiguity and underlying associations readers may have for the words chosen.
\\\\
\textbf{\textit{Encoding schemes}} - This is the physical method of encoding used to represent the fingerprint. For example, ``Hexadecimal'' or ``Words'' are a unique way to represent a fingerprint and thus, are encoding schemes.
\\\\
\textbf{\textit{Method of comparison}} - This is how the user assesses the encoding scheme. The most common example and the one used as default is ``Compare-and-Confirm'' (CaC). CaC, as the name suggests, is the \textit{comparison} of two fingerprints on different devices or mediums that are then \textit{confirmed}. The following sections explain alternative methods of comparisons.

\section{Literature Review}
This section will now discuss the current research around the proposed topic.

The section is organized as follows:

\begin{itemize}
    \item \textbf{Authentication ceremony performance} \\
    This section explains the current research on performance of various encoding schemes and methods of comparison. This is then complemented with an assessment of the experimentation design for the respective studies.

    \item \textbf{Encoding schemes} \\
    The following section then discusses research around the design of an encoding scheme. Focus in this section will be allocated to the technical details around the creation and design of relevant schemes.

    \item \textbf{Attacks on encoding schemes} \\
    The final section will then discuss literature that investigates the creation of attacks against encoding schemes in order to deceive the respective users.
\end{itemize}

@@

\subsubsection*{Encoding scheme performance}
Results 
from the literature consistently show the effectiveness of language-based encodings such as Words or Sentences with accuracies ranging from 94\% \cite{dechand2016empirical}\cite{tan2017can}\cite{kainda2009usability}. In all cases, these were the best schemes from the sets assessed. The exception to this is the work performed by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study} in 2009 with Words achieving an abnormal accuracy of 63.00\%.
\\\\
Aside from textual representations were graphical schemes. Examples of schemes assessed were: Random Art\cite{perrig1999hash}, Flag\cite{ellison2003public}, T-Flag\cite{lin2010spate}, Vash\footnote{https://github.com/thevash/vash}, OpenSSH visual host key and Unicorns\footnote{https://unicornify.pictures/}, among others. These schemes had mixed accuracy with ranges as large as 50\% - 94\% in work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study}. The only other paper assessing graphical representations was the work of \textbf{Tan \textit{et al.}}\cite{tan2017can} where they also achieved mixed results with accuracies ranging from 46\% to 90\%.
\\
\begin{table}[h!]
    \makebox[\textwidth][c]{
        \input{tables/fingerprint_times.tex}
    }%
    \caption{Timing results in seconds for the related schemes}
    \label{tab:times}
\end{table}
\\
In terms of usability of graphical schemes, the literature concurred on their high usability. The comparison speed of these schemes were all among the quickest (See Table \ref{tab:times} for an overview of timings). Other work also indirectly concurred with graphical encodings having significantly quicker comparison times compared to non-graphical schemes \cite{dechand2016empirical}\cite{kainda2009usability}.
In terms of research into the performance of graphical schemes, the literature does not contain an extensive review with literature only containing two papers. There is also no overlap in the schemes assessed with each paper reviewing a unique set. This is, therefore, a promising candidate for further research.

@@

The findings from the paper showed a high false negative rate for all the schemes; this is a consideration also missing from the literature. Alongside this, results for usability were lower in a remote setting for all the schemes. The author, however, comments on the expected nature of this result.\\
Images were assessed as being the most secure method of authentication in the remote setting but voted as the method with the worst usability. This conclusion is highly inconsistent with all other work in this area. However, this could be due to the unique setting of remote verification, resulting in distinct results from user studies. Without further work in this area, it is difficult to validate these results conclusively.

Alongside this, it was shown in work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study} that age and gender do not affect the accuracy of the scheme. However, younger participants were considerably faster. Furthermore, findings also showed that language comprehension helped in discerning small differences between schemes encoded in Chinese, Japanese, Korean or English. Subsequently, knowledge of the language did not assist in differentiating more significant changes in the schemes as these had high accuracy regardless. These were interesting and unique considerations. Further work could aim to corroborate these conclusions.

@@

\subsubsection*{Methods of comparison performance}
Aside 
from ``Compare-and-Confirm'' (CaC), there is ``Compare-and-Select'' (CaS) and ``Compare-and-Enter'' (CaE). CaS is the method where one device displays the fingerprint, and the other user is provided with several options. The user then has to choose the correct value from the list of candidates. If there is no match, the user must deny the connection attempt. The creation of CaS was due to concerns that CaC would be ``too easy'' for users leading to complacency and errors\cite{uzun2007usability}. CaE is designed for scenarios where both devices might not have a display, i.e. pairing between a phone and a keyboard. One device displays the checksum. This checksum is entered into the other device. The first device then compares the entered string and checks for a match.

Research has been performed assessing the performance of these schemes and how they affect the ultimate security of the authentication ceremony. The literature agrees on CaC being the best overall scheme to compare fingerprints \cite{tan2017can}\cite{uzun2007usability} with CaS being highlighted for its poor security and usability. CaE has  had contrasting results. \cite{uzun2007usability} discarded it after one round due to ``poor usability.'' However, \cite{tan2017can} considered it the best method overall for usability and security. These conflicting results, therefore, show polarisation in the results of CaE. However, this could be due to the different overall use-cases of the studies. Validation of results from either study would, therefore, be an area of additional study.

% One aspect of note from the work by \textbf{Ersin Uzun \textit{et al.}}\cite{uzun2007usability} is their studies context  of  “Secure  device  pairing”  and  thus  some comparison schemes are tested that are not relevant in a fingerprint comparison context, for example “Choose-and-Enter” (A participant chooses a passphrase from an available list and enters it into the other device).  Thus, only relevant comparison schemes and their results were extracted and compared. \\

\subsubsection*{Experimentation methodology comparison}
To further look into the validity of previously discussed results, it is necessary to asses how the respective studies reached these conclusions. Areas for consideration are scheme entropy, attacker strength and participant attributes.

The starkest limitations of the literature are the range of participants and encoding entropy. The worst studies tested only 22-bits of entropy. This makes it difficult to compare results directly. One of the papers this most effects is the early work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study}  where their highest entropy is 28-bits. This level of entropy was inadequate even at the time of publication. There is an attempt by the authors' to address this issue in the later stages of the paper where they write \textit{``[...] increasing entropy is not a solution because it sacrifices usability and accuracy. With more entropy, representations will contain longer sequences of characters or more minute details which will lead to increased time and errors during comparisons.''}. This statement is backed up with no empirical evidence, and the authors' fail to consider how the low entropy would affect the overall security of the schemes.

@@

Hexadecimal and numerical schemes were not included as encoding schemes in work by \textbf{Hsiao, \textit{et al.}}\cite{hsiao2009study} (some of the most widespread encoding schemes). This decision was based on their claims on similarities to Base32 and ``well-known deficiencies" in the excluded schemes. This point was provided with no further justification or quantified in any way. It is also not consistent with available research, for example, in \textit{``Empirical Study of Textual Key-Fingerprint Representations''}\cite{dechand2016empirical} it was shown numerical representations performed significantly better than that of Base32.

>>>>

The chapter contains the following sections:

\begin{itemize}
    \item \textbf{Project context} \\ 
    Provides the user with the base knowledge required to understand the context and purpose of the following literature review.

    \item \textbf{Definition of terms}\\
    Specific definitions relevant to the literature review are initially defined.

    \item \textbf{Literature Review} \\
    A review of currently available literature on relevant topics. Recommendations and research gaps are identified and discussed.

@@

To fully understand the literature review, it is necessary to introduce and explain the base knowledge before continuing.

@@

Asymmetric cryptography facilitates the secure encryption of messages in end-to-end encryption (E2EE), verification of digital signatures, and sharing of pre-communication secrets, among others.

The asymmetry stems from the use of "Public" and "Private" keys. The public-key is used to encrypt data that only the respective private-key can decrypt. Hence, this means that sharing the keys required to encrypt can be performed across insecure channels.

An E2EE connection uses the key types discussed previously to exchange messages between verified parties securely. However, this hinges on the verification of the initial parties, if one party impersonates another and receives sensitive communication, the secure encryption is completely circumvented. Therefore, the correct identification of parties is crucial to maintaining security. The exploitation of this is known as a Man-in-the-middle (MiTM) attack. This attack is commonly bidirectional, and if executed correctly, there is often no discerning change to the user experience. Figure \ref{fig:mitm} contains a visual representation of this attack.

@@

Due to the assumption that the attacking party (Eve) does not have the same public-private key pair as either party (Alice or Bob), the fingerprint of the public keys are, therefore, different, and, thus can be used for identification. There are, however, issues in how a user should link a key's fingerprint to a real-world entity. This paper will, however, assume that users have access to the respective parties' correct fingerprint, as the protocols used to achieve this are outside the project's scope.

The fingerprint of the key is generated by running the main key components through a secure one-way hash function such as SHA-1. This process produces a digest of a fixed length that can be used to compare keys. Therefore, the comparison of expected and actual fingerprints can be used to detect MiTM attacks.
Historically these have been represented as a hexadecimal string whereon verification fingerprints are compared between two substrates, for example, a monitor screen and a business card.
This process is known as the \textit{``authentication ceremony.''} System designers are, however, moving away from this structure and are now implementing a combined key. This change is due to the assumed increase in usability. Users now compare a single fingerprint, instead of two sets separately. In the case of WhatsApp for example, the connection fingerprint of two parties keys is the first 30 bytes of an SHA-512\footnote{5200 iterations} hash of each parties' identity key, these are then both concatenated together to form a single key\cite{whatsapp2017paper}. This process produces a unique-key communication pair.
\\\\
Due to the manual comparison of the fingerprint, length, and format is a core design consideration. Prior research has shown that the average human can only hold around 7-digits worth of data in their working memory\cite{miller1956magical}. This rules out the possibility of comparing complete digests. For example, SHA-1 is 40 hex digits (160-bit) and, therefore, difficult to effectively compare. Therefore, if human interaction is required; there is a need for schemes that work effectively with consideration to individual limitations.

Research in this area has taken various encoding schemes and compared their fallibility to impersonated attacks. The main elements of comparison have been the ``accuracy of attack detection" and ``time to compare." In this paper, these are considered the metrics of ``security'' and ``usability,'' respectively.

\section{Definition of terms}
Throughout the paper, various terms are used to define elements of the authentication ceremony relevant to fingerprints verification. Explicit descriptions are, therefore, stated in this section. These are used to remove ambiguity and underlying associations readers may have for the words chosen.
\\\\
\textbf{\textit{Encoding schemes}} - are the physical method of encoding used to represent the fingerprint. For example, ``Hexadecimal'' or ``Words'' are a unique way to represent a fingerprint and thus, are encoding schemes.
\\\\
\textbf{\textit{Method of comparison}} - is the way a user assesses the encoding scheme. The most common example and the one used as default is ``Compare-and-Confirm'' (CaC). CaC, as the name suggests, is the \textit{comparison} of two fingerprints on different devices or mediums that are then \textit{confirmed}. The following sections explain alternative methods of comparisons.

\section{Literature Review}
This section discusses the current research around the proposed topic.

The organization of the section is as follows:

\begin{itemize}
    \item \textbf{Authentication ceremony performance} \\
    This section explains the current research on the performance of various encoding schemes and methods of comparison, alongside an assessment of the literature's experimental design.

    \item \textbf{Encoding schemes} \\
    The following section then discusses research around the design of an encoding scheme. The focus of the section is the technical details regarding the creation and design of schemes.

    \item \textbf{Attacks on encoding schemes} \\
    The final section discusses literature that investigates the creation of actual attacks against encoding schemes.
\end{itemize}

@@

\subsubsection*{Encoding scheme performance}
Results 
from the literature consistently show the effectiveness of language-based encodings such as Words or Sentences with accuracies ranging up from 94\% \cite{dechand2016empirical}\cite{tan2017can}\cite{kainda2009usability}. In all cases, these were the best schemes from the sets assessed. The exception to this is the work performed by \textbf{Hsiao \textit{et al.}}\cite{hsiao2009study} in 2009 with Words achieving an abnormal accuracy of 63\%.
\\\\
Aside from textual representations were graphical schemes. Examples of schemes assessed were: Random Art\cite{perrig1999hash}, Flag\cite{ellison2003public}, T-Flag\cite{lin2010spate}, Vash\footnote{https://github.com/thevash/vash}, OpenSSH visual host key and Unicorns\footnote{https://unicornify.pictures/}, among others. These schemes had mixed accuracy with ranges as large as 50\% - 94\% in work by \textbf{Hsiao \textit{et al.}}\cite{hsiao2009study}. The only other paper assessing graphical representations was the work of \textbf{Tan \textit{et al.}}\cite{tan2017can} where they also achieved mixed results with accuracies ranging from 46\% to 90\%.
\\
\begin{table}[h!]
    \makebox[\textwidth][c]{
        \input{tables/fingerprint_times.tex}
    }%
    \caption{Timing results in seconds for the related schemes}
    \label{tab:times}
\end{table}
\\
In terms of usability of graphical schemes, the literature concurred on their high usability. The comparison speed of these schemes were all among the quickest (See Table \ref{tab:times} for an overview of timings). Other work also indirectly agreed with graphical encodings having significantly quicker comparison times compared to non-graphical schemes \cite{dechand2016empirical}\cite{kainda2009usability}.
In terms of research into the performance of graphical schemes, the literature does not contain an extensive review with literature only containing two papers. There is also no overlap in the schemes assessed with each paper reviewing a unique set. This area is, therefore, a promising candidate for further research.

@@

The findings from the paper showed a high false-negative rate for all the schemes. This consideration is also missing from the literature. Alongside this, results for usability were lower in a remote setting for all the schemes. The author, however, comments on the expected nature of this result.\\
Images were identified as being the most secure method of authentication in the remote setting but voted as the method with the worst usability. This conclusion is highly inconsistent with all other work in this area. However, this could be due to the unique setting of remote verification, resulting in distinct results from user studies. Without further work in this area, it is difficult to validate these results conclusively.

Alongside this, it was shown in work by \textbf{Hsiao\textit{et al.}}\cite{hsiao2009study} that age and gender do not affect the accuracy of the scheme. However, younger participants were considerably faster. Furthermore, findings also showed that language comprehension helped in discerning small differences between schemes encoded in Chinese, Japanese, Korean, or English. Subsequently, knowledge of the language did not assist in differentiating more significant changes in the schemes as these had high accuracy regardless. These were unique considerations. Further work could aim to corroborate these conclusions.

@@

\subsubsection*{Methods of comparison performance}
Aside 
from ``Compare-and-Confirm'' (CaC), there is ``Compare-and-Select'' (CaS) and ``Compare-and-Enter'' (CaE). CaS is the method where one device displays the fingerprint, and the other user is provided with several options. The user then has to choose the correct value from the list of candidates. If there is no match, the user must deny the connection attempt. The creation of CaS was due to concerns that CaC would be ``too easy'' for users leading to complacency and errors\cite{uzun2007usability}. The design of CaE is for scenarios where both devices might not have a display, i.e., pairing between a phone and a keyboard. One device displays the checksum. This checksum is then entered into the other device. The first device then compares the entered string and checks for a match.

Research has assessed the performance of these schemes and how they affect the security of the authentication ceremony. The literature agrees on CaC being the best overall scheme to compare fingerprints \cite{tan2017can}\cite{uzun2007usability} with CaS highlighted for its poor security and usability. CaE has had conflicting results. \cite{uzun2007usability} discarded it after one round due to ``poor usability.'' However, \cite{tan2017can} considered it the best method overall for usability and security. These conflicting results, therefore, show polarisation in the results of CaE. However, this could be due to the different overall use-cases of the studies. Validation of results from either study would, therefore, be an area of additional study.

\subsubsection*{Experimentation methodology comparison}
To further look into the validity of previously discussed results, it is necessary to asses how the respective studies reached their conclusions. Areas for consideration are scheme entropy, attacker strength, and participant demographics.

The starkest limitations of the literature are the range of participants and encoding entropy. The worst studies tested only 22-bits of entropy. This lack of entropy makes it challenging to compare results directly. One of the papers this most effects is the early work by \textbf{Hsiao \textit{et al.}}\cite{hsiao2009study}  where their highest entropy is 28-bits. This level of entropy was inadequate even at the time of publication. There is an attempt by the authors' to address this issue in the later stages of the paper where they write \textit{``[...] increasing entropy is not a solution because it sacrifices usability and accuracy. With more entropy, representations will contain longer sequences of characters or more minute details which will lead to increased time and errors during comparisons.''}. This statement is backed up with no empirical evidence, and the authors fail to consider how the low entropy would affect the overall security of the schemes.

@@

Hexadecimal and numerical schemes were not included as encoding schemes in work by \textbf{Hsiao \textit{et al.}}\cite{hsiao2009study} (some of the most widespread encoding schemes). This decision was based on their claims on similarities to Base32 and ``well-known deficiencies" in the excluded schemes. This point was provided with no further justification or quantified in any way. It is also not consistent with available research, for example, in \textit{``Empirical Study of Textual Key-Fingerprint Representations''}\cite{dechand2016empirical} it was shown numerical representations performed significantly better than that of Base32.
